Fractional Types

### 2016-12-18 typespp.tex
forall is dual to exists

The looping groupoid $\looping{G}$ has the elements $g$ of the group $G$ as
objects and morphisms between elements in the same conjugacy class; that is,
there is a
morphism between elements $g$ and $h$ iff there exists a $x \in G$ such that
$g = x^{-1} \cdot g \cdot x$. Fixing a type $\tau$ and a particular 1-combinator
$p : \tau \iso \tau$, we get a group whose elements are the iterates $p^k$. The
corresponding looping groupoid has all the iterates as objects, and morphisms
between $p^{k_1}$ and $p^{k_2}$ whenever there exists an integer $j$ such that
$p^{k_2}$ is equivalent to $p^{-j} \odot p^{k_1} \odot p^{j}$. By
Lem.~\ref{lem:distiterplus}, the latter term is $\isotwo$-equivalent to
$p^{-j+k_1+j}$ which is $\isotwo$-equivalent to $p^{k_1}$. In other words, we
have a morphism between $p^{k_1}$ and $p^{k_2}$ whenever they are
$\isotwo$-equivalent which is consistent with our definition of iteration
groupoids.

@@

If we are to build a compositional programming language around iteration and
symmetry groupoids, we need to ensure that they compose sensibly with the
existing type formers. Groupoids, viewed as categories, come associated with
natural notions of sums and products, and one might expect or hope that
identities which hold for rational numbers lift to identities in our
situation. The most prominent and important identity we seek is:
\[
\order{p} \circledast \iorder{p} \simeq \order{\idiso}
\]
The left hand side is the categorical product of the iteration groupoid and
symmetry groupoid for some arbitrary 1-combinator $p$. The cardinality of this
groupoid is 1. The right hand side is the iteration groupoid for the identity
1-combinator which also has cardinality 1. The groupoids on either side are
however not categorically equivalent and hence $\simeq$ cannot be categorical
equivalence.  There are however other notions of groupoid equivalence, such as
Morita equivalence~\cite[C5.3]{johnstone2002sketches}~\cite{looping}, that
identify the two groupoids above.  Although the notion of equivalence we use in
the next section appears novel, it is related ``in spirit'' to Morita
equivalence and we find it useful to give a small example. We will verify that
the product of the groupoid $\order{\permfive}$ of cardinality 3 and the
groupoid $\iorder{\permfive}$ of cardinality $\frac{1}{3}$ is equivalent
to the trivial
groupoid with only one distinct object and one trivial automorphism. To avoid
clutter we assume the groupoids are strict, e.g., that the groupoid
$\order{\permfive}$ has three elements (instead of 3 isomorphism classes of
elements). The definition of \emph{Morita equivalence} in this case reduces to
to finding a \emph{Morita morphism} from $\order{\idiso}$ to
$\order{\permfive} \circledast \iorder{\permfive}$. This is a functor $\psi$
from the unit groupoid to the product groupoid, say $G$, with an additional
condition that the commutative diagram in figure~\ref{fig:morita} is a
\emph{pullback}, where $s,t$ are the \emph{source} and \emph{target} maps,
and $\mathbb{1}$, $\mathbb{3}$ are the $1$- and $3$-point discrete groupoids
defined previously.
The intuition is that the codomain consists of three objects
with associated morphisms, each of them counting as a ``third.'' The functor
from $\order{\idiso}$ to $\order{\permfive} \circledast \iorder{\permfive}$ must
include an arbitrary choice of ``which third'' to target and the universality
condition of the pullback ensures that the choice is ultimately insignificant as
it leads to the same result as every other possible functor. To summarize the
equivalence between the two groupoids requires a \emph{local} map that chooses a
target and a \emph{global} condition ensuring that the choice ultimately
annihilates. The challenge we are facing is to turn this idea into an operational
semantics that somehow combines the local and global aspects of the
equivalence. We will be inspired by the folklore result that ``substitution into a
dependent type is a pullback''~\cite{subpullback} (but which turns out to have
been fully explained in~\cite{practicalfoundations}) as explained in the next
section.

\begin{figure}[h]
  \centering
  \begin{tikzcd}
    Q
    \arrow[drr, bend left, "{\xi_1}"]
    \arrow[ddr, bend right, "{(s,t)}"]
    \arrow[dr, dotted] & & \\
    & \order{\idiso} \arrow[r, "{\psi_1}"] \arrow[d, "{(s,t)}"]
    & G \arrow[d, "{(s,t)}"] \\
    & \mathbb{1} \times \mathbb{1} \arrow[r, "{\psi_0 \times \psi_0}"]
    & \mathbb{3} \times \mathbb{3}
  \end{tikzcd}
  \caption{
    \label{fig:morita}
    Pullback diagram for a Morita morphism,
    $\order{\idiso} \rightarrow \order{\permfive} \circledast
    \iorder{\permfive}$
  }
\end{figure}

@@

We are now ready to turn the groupoid constructions from the previous section
into an extension of $\Pi$ with new type constructors and combinators for
creating and manipulating syntactic counterparts to $\order{p}$ and
$\iorder{p}$.  Just adding the ``obvious'' types and combinators does
not work; however, as the failures are quite informative, we nevertheless go
through this first attempt.  This will help illustrate why we need extra
machinery in Sec.~\ref{sec:deppointedtypes}.

@@

%%%%%
\subsection{First attempt}

@@

The syntactic category $\ft$ of fractional types introduces type expressions
for iteration groupoids, symmetry groupoids, and their products. The
combinators that relate these fractional types are in Fig.~\ref{pifrac:comb}.
The bottom group includes identity and sequential composition~$\circledcirc$
combinators and ensures that the combinators can be applied anywhere inside a
product $\circledast$. There is also a combinator that relates two types
$\order{c_1}$ and $\order{c_2}$ via $\order{\alpha}$ whenever $\alpha : c_1
\isotwo c_2$.  Indeed if $c_1$ and $c_2$ are considered equivalent via a
$2$-combinator then we also consider their iteration groupoids to be equivalent.
The top group ensures that $\circledast$ carries a symmetric monoidal structure
with combinators witnessing the unit, commutativity, and associativity
properties. The last two pairs of combinators $\eta\!- / \epsilon-$ and
$\eta\!+ / \epsilon+$ are inspired from the definition of compact closed
categories~\cite{ccc} which are symmetric monoidal categories in which every
object has a dual. They are the ones that witness the ``fractional'' nature of
symmetry groupoids as motivated in Sec.~\ref{sub:products}.

The last new syntactic category in $\Pi^/$ is that of values which deserves some
additional discussion. When types denote sets, values of a type are clear:
they are just the elements of the set.  In the case when types are groupoids,
this is less clear, especially when the groupoid cardinality is a proper
fraction.  What we get instead are equivalence classes of values.
The idea is not uncommon: in the conventional $\lambda$-calculus, we list
$\lambda x.x$ and $\lambda y.y$ as separate values of type
$\tau \rightarrow \tau$ and then provide a separate equivalence relation
($\alpha$-equivalence) to express the fact that these two values are
indistinguishable. The treatment in our setting is similar: for iteration
groupoids $\order{c}$, every distinct iterate $c^k$ is a listed as a separate
value with the understanding that some iterates are
$\isotwo$-equivalent. Similarly for symmetry groupoids $\iorder{c}$, every
morphism is listed as a value $1/c^k$ with the same understanding that that some
of these morphisms are $\isotwo$-equivalent.

@@

%%%%%
\subsection{Dependent Pointed Types}\label{sec:deppointedtypes}

@@

\begin{definition}[Dependent $\Pi^/$]
\label{def:pifracd}
The syntax of dependent $\Pi^/$ differs from Def.~\ref{def:pifrac} as follows:
\[\begin{array}{lrcl}
(\textrm{Fractional Types}) & \mathbb{T_\bullet} &::=&
  [\order{c},c^k] \alt
  [\iorder{c},1/c^k] \alt
  [\mathbb{T}_1 \circledast \mathbb{T}_2,\langle v_1,v_2 \rangle] \alt
  \forall k.\mathbb{T_\bullet} \\
 (\textrm{/-combinators}) & \rho_\bullet &:&
  \mathbb{T}_{\bullet 1} \leftrightblackspoon \mathbb{T}_{\bullet 2} ~::=
  [\textit{see Fig.~\ref{pifracd:comb}}]
\end{array}\]
\end{definition}

\begin{figure}[t]
{\footnotesize
\[\begin{array}{rrcll}
\unitetl/ :&  \forall\overline{k}.[\order{\idiso} \circledast \mathbb{T},
                                                   \langle \idiso,v \rangle]
  & \leftrightblackspoon &
                    \forall\overline{k}.[\mathbb{T},v]
  &: \unititl/ \\
\unitetr/ :&  \forall\overline{k}.[\mathbb{T} \circledast \order{\idiso},
                                                    \langle v,\idiso]
  & \leftrightblackspoon &
                    \forall\overline{k}.[\mathbb{T},v]
  &: \unititr/ \\
\swapt/ :&  \forall\overline{k}.[\mathbb{T}_1 \circledast \mathbb{T}_2,
                                                   \langle v_1,v_2 \rangle]
 & \leftrightblackspoon &
                   \forall\overline{k}.[\mathbb{T}_2 \circledast \mathbb{T}_1,
                                                  \langle v_2,v_1 \rangle]
  &: \swapt/ \\
  \assoclt/ :&  \forall\overline{k}.[\mathbb{T}_1 \circledast (\mathbb{T}_2 \circledast \mathbb{T}_3),
                                        \langle v_1,\langle v_2,v_3 \rangle]
  & \leftrightblackspoon &
                       \forall\overline{k}.[(\mathbb{T}_1 \circledast \mathbb{T}_2) \circledast \mathbb{T}_3,
                                       \langle \langle v_1,v_2 \rangle , v_3 \rangle]
  &: \assocrt/ \\
\\
\eta- :& [\order{\idiso},\idiso] & \leftrightblackspoon &
  \forall{k}.[\iorder{c} \circledast \order{c}, \langle 1/c^k, c^k \rangle] &: \epsilon- \\
\eta+ :& [\order{\idiso},\idiso] & \leftrightblackspoon &
  \forall{k}.[\order{c} \circledast \iorder{c}, \langle c^k,1/c^k \rangle] &: \epsilon+ \\
\end{array}\]
\[\begin{array}{c}
\Rule{}
{}
{\jdg{}{}{\idiso/ : \forall\overline{k}.[\mathbb{T},v] \leftrightblackspoon
  \forall\overline{k}.[\mathbb{T},v]}}
{}
~~
\Rule{}
{\jdg{}{}{\alpha : c_1 \isotwo c_2}}
{\jdg{}{}{(\order{\alpha})_k :
 [\order{c_1},c_1^k]  \leftrightblackspoon
 [\order{c_2}.c_2^k]}}
{}
\\
\\
\Rule{}
{\jdg{}{}{\rho_1 : \forall{\overline{k_1}}.
  [\mathbb{T}_1,v_1] \leftrightblackspoon [\mathbb{T}_2,v_2]} \quad
  \rho_2 : \forall{\overline{k_2}}.
  [\mathbb{T}_2,v_2] \leftrightblackspoon [\mathbb{T}_3,v_3]}
{\jdg{}{}{\rho_1 \circledcirc \rho_2 :
 \forall{\overline{k_1}\overline{k_2}}.
 [\mathbb{T}_1,v_1] \leftrightblackspoon [\mathbb{T}_3},v_3]}
{}
\\
\\
\Rule{}
{\jdg{}{}{\rho_1 :
  \forall{\overline{k_1}}.[\mathbb{T}_1,v_1] \leftrightblackspoon
  \forall{\overline{k_2}}.[\mathbb{T}_2,v_2]}
 \qquad \rho_2 :
  \forall{\overline{k_3}}.[\mathbb{T}_3,v_3] \leftrightblackspoon
  \forall{\overline{k_4}}.[\mathbb{T}_4,v_4]}
{\jdg{}{}{\rho_1 \circledast \rho_2 :
  \forall{\overline{k_1k_2k_3k_4}}.
  [\mathbb{T}_1 \circledast \mathbb{T}_3, \langle v_1,v_3 \rangle]
  \leftrightblackspoon
  [\mathbb{T}_2 \circledast \mathbb{T}_4}, \langle v_2,v_4 \rangle]}
{}
\\
\\
\Rule{}
{\jdg{}{}{}}
{\jdg{}{}{\synch_i : \forall{\overline{k}k'}.[\mathbb{T},v]
  \leftrightblackspoon
   \forall{\overline{k}}.[\mathbb{T},v[k'\mapsto i]]}}
{}
\end{array}\]
Each combinator $\rho_\bullet$ has an inverse.
}
\caption{$\Pi^/$ fractional combinators (dependent version)}
\label{pifracd:comb}
\end{figure}

The first step is to enrich the types to pointed types $\mathbb{T_\bullet}$
where each type $\mathbb{T}$ refers to a particular value $v$ in
$\mathbb{T}$. We then allow indices of values $k$ to be range over a collection
of types. In particular, the type
$\forall k.[\order{c} \circledast \iorder{c}, \langle c^k, 1/c^k \rangle]$
imposes a constraint on the two values associated with the types $\order{c}$ and
$\iorder{c}$; they must be ``synchronized'' sharing the same $k$. In general, a
pointed type will be of the form $\forall\overline{k}.[\mathbb{T},v]$ for some
sequence of integers $\overline{k}$ and some value $v$ in $\mathbb{T}$ which
might of course refer to the integers $\overline{k}$ to enforce some
synchronization among its sub-values.

Here is how the circuit in Fig.~\ref{fig:zigzag} evaluates in the forward and
backwards directions:
\[\begin{array}{rcl}
[\order{c},c^i]
&\stackrel{\unitetl/}{\longmapsto}&
  [\order{id}\circledast\order{c}, \langle \idiso,c^i \rangle] \\
&\stackrel{\eta+\circledast\idiso/}{\longmapsto}&
  \forall{k}.[(\order{c}\circledast\iorder{c})\circledast\order{c},
  \langle \langle c^k,1/c^k \rangle,c^i \rangle] \\
&\stackrel{\assocrt/}{\longmapsto}&
  \forall{k}.[\order{c}\circledast(\iorder{c}\circledast\order{c}),
  \langle c^k, \langle 1/c^k,c^i \rangle\rangle] \\
&\stackrel{\synch_i}{\longmapsto}&
  [\order{c}\circledast(\iorder{c}\circledast\order{c}),
  \langle c^i, \langle 1/c^i,c^i \rangle\rangle] \\
&\stackrel{\idiso/\circledast\epsilon-}{\longmapsto}&
  [\order{c}\circledast\order{\idiso},
  \langle c^i, \idiso \rangle] \\
&\stackrel{\unitetr/}{\longmapsto}&
  [\order{c}, c^i]
\end{array}\]

\[\begin{array}{rcl}
  [\order{c}, c^i]
&\stackrel{\unitetr/}{\longmapsfrom}&
  [\order{c}\circledast\order{\idiso},
  \langle c^i, \idiso \rangle] \\
&\stackrel{\idiso/\circledast\epsilon-}{\longmapsfrom}&
  \forall{k}.[\order{c}\circledast(\iorder{c}\circledast\order{c}),
  \langle c^i, \langle 1/c^k,c^k \rangle\rangle] \\
&\stackrel{\synch_i}{\longmapsfrom}&
 [\order{c}\circledast(\iorder{c}\circledast\order{c}),
  \langle c^i, \langle 1/c^i,c^i \rangle\rangle] \\
&\stackrel{\assocrt/}{\longmapsfrom}&
  [(\order{c}\circledast\iorder{c})\circledast\order{c},
  \langle \langle c^i,1/c^i \rangle,c^i \rangle] \\
&\stackrel{\eta+\circledast\idiso/}{\longmapsfrom}&
  [\order{c}\circledast\order{\idiso}, \langle c^i, \idiso] \\
&\stackrel{\unitetl/}{\longmapsfrom}&
  [\order{c},c^i]
\end{array}\]

>>>>

The looping groupoid $\looping{G}$ has the group elements as objects and
morphisms between elements in the same conjugacy class, that is, there is a
morphism between elements $g$ and $h$ iff there exists a $x \in G$ such that
$g = x^{-1} \cdot g \cdot x$. Fixing a type $\tau$ and a particular 1-combinator
$p : \tau \iso \tau$, we get a group whose elements are the iterates $p^k$. The
corresponding looping groupoid has all the iterates as objects and morphisms
between $p^{k_1}$ and $p^{k_2}$ whenever there exists an integer $j$ such that
$p^{k_2}$ is equivalent to $p^{-j} \odot p^{k_1} \odot p^{j}$. By
Lem.~\ref{lem:distiterplus}, the latter term is $\isotwo$-equivalent to
$p^{-j+k_1+j}$ which is $\isotwo$-equivalent to $p^{k_1}$. In other words, we
have a morphism between $p^{k_1}$ and $p^{k_2}$ whenever they are
$\isotwo$-equivalent which is consistent with our definition of iteration
groupoids.

@@

If we are to build a compositional programming language around iteration and
symmetry groupoids, we need to ensure that they compose sensibly with the
existing type formers. Groupoids, viewed as categories, come associated with
natural notions of sums and products, and one might expect or hope that
identities which hold for rational numbers lift to identities in our
situation. The most prominent and important identity we seek is:
\[
\order{p} \circledast \iorder{p} \simeq \order{\idiso}
\]
The left hand side is the categorical product of the iteration groupoid and
symmetry groupoid for some arbitrary 1-combinator $p$. The cardinality of this
groupoid is 1. The right hand side is the iteration groupoid for the identity
1-combinator which also has cardinality 1. The groupoids on either side are
however not categorically equivalent and hence $\simeq$ cannot be categorical
equivalence.  There are however other notions of groupoid equivalence, such as
Morita equivalence~\cite[C5.3]{johnstone2002sketches}~\cite{looping}, that
identify the two groupoids above.  Although the notion of equivalence we use in
the next section appears novel, it is related ``in spirit'' to Morita
equivalence and we find it useful to give a small example. We will verify that
the product of the groupoid $\order{\permfive}$ of cardinality 3 and the
groupoid $\iorder{\permfive}$ of cardinality $\frac{1}{3}$ is the trivial
groupoid with only one distinct object and one trivial automorphism. To avoid
clutter we assume the groupoids are strict, e.g., that the groupoid
$\order{\permfive}$ has three elements (instead of 3 isomorphism classes of
elements). The definition of \emph{Morita equivalence} in this case reduces to
to finding a \emph{Morita morphism} from $\order{\idiso}$ to
$\order{\permfive} \circledast \iorder{\permfive}$. This is a functor $\psi$
from the unit groupoid to the product groupoid, say $G$, with an additional
condition that the commutative diagram in figure~\ref{fig:morita} is a
\emph{pullback}. The intuition is that the codomain consists of three objects
with associated morphisms, each of them counting as a ``third.'' The functor
from $\order{\idiso}$ to $\order{\permfive} \circledast \iorder{\permfive}$ must
include an arbitrary choice of ``which third'' to target and the universality
condition of the pullback ensures that the choice is ultimately insignificant as
it leads to the same result as every other possible functor. To summarize the
equivalence between the two groupoids requires a \emph{local} map that chooses a
target and a \emph{global} condition ensuring that the choice ultimately
annihilates. The challenge we are facing is to turn this idea into an operational
semantics that somehow combines the local and global aspects of the
equivalence. We will be inspired by the folklore result that ``substitution into a
dependent type is a pullback''~\cite{subpullback} as explained in the next section.

\begin{figure}[h]
  \centering
  \begin{tikzcd}
    Q
    \arrow[drr, bend left, "{\xi_1}"]
    \arrow[ddr, bend right, "{(s,t)}"]
    \arrow[dr, dotted] & & \\
    & \set{\idiso} \arrow[r, "{\psi_1}"] \arrow[d, "{(s,t)}"]
    & G_1 \arrow[d, "{(s,t)}"] \\
    & \set{\ast} \times \set{\ast} \arrow[r, "{\psi_0 \times \psi_0}"]
    & G_0 \times G_0
  \end{tikzcd}
  \caption{
    \label{fig:morita}
    Pullback diagram for a Morita morphism,
    $\order{\idiso} \rightarrow \order{\permfive} \circledast
    \iorder{\permfive}$
  }
\end{figure}

@@

We are now ready to turn the groupoid constructions from the previous section into an extension of $\Pi$ with new type constructors and combinators for creating and manipulating syntactic counterparts to $\order{p}$ and $\iorder{p}$.

@@

%%%%%
\subsection{Non-Dependent Version}

@@

The syntactic category $\ft$ of fractional types introduces type expressions for
iteration groupoids, symmetry groupoids, and their products. The combinators
that relate these fractional types are in Fig.~\ref{pifrac:comb}. The bottom
group includes identity and sequential composition~$\circledcirc$ combinators
and ensures that the combinators can be applied anywhere inside a product
$\circledast$. There is also a combinator that relates two types $\order{c_1}$
and $\order{c_2}$ via $\order{\alpha}$ whenever $\alpha : c_1 \isotwo
c_2$.
Indeed if $c_1$ and $c_2$ are considered equivalent via a 2-combinator then we
also consider their iteration groupoids to be equivalent. The top group ensures
that $\circledast$ carries a symmetric monoidal structure with combinators
witnessing the unit, commutativity, and associativity properties. The last two
pairs combinators $\eta-/\epsilon-$ and $\eta+/\epsilon+$ are inspired from the
definition of compact closed categories~\cite{ccc} which are symmetric monoidal
categories in which every object has a dual. They are the ones that witness the
``fractional'' nature of symmetry groupoids as motivated in Sec.~\ref{sub:products}.

The last new syntactic category in $\Pi^/$ is that of values which deserves some
background discussion. When the types denote sets, it is evident what it means
to have a value of a given type: it is just an element of the set. When types
denote groupoids, it is less clear what it means to have a value, especially
when the total number of values, as reported by the groupoid cardinality, is a
proper fraction. In that case we the values up to an equivalence relation. The
idea is not uncommon: in the conventional $\lambda$-calculus, we list
$\lambda x.x$ and $\lambda y.y$ as separate values of type
$\tau \rightarrow \tau$ and then provide a separate equivalence relation
($\alpha$-equivalence) to express the fact that these two values are
indistinguishable. The treatment in our setting is similar: for iteration
groupoids $\order{c}$, every distinct iterate $c^k$ is a listed as a separate
value with the understanding that some iterates are
$\isotwo$-equivalent. Similarly for symmetry groupoids $\iorder{c}$, every
morphism is listed as a value $1/c^k$ with the same understanding that that some
of these morphisms are $\isotwo$-equivalent.

@@

\begin{figure}[t]
{\footnotesize
\[\begin{array}{rrcll}
\unitetl/ :& [\order{\idiso} \circledast \mathbb{T},
                                                   \langle \idiso,v \rangle]
  & \leftrightblackspoon &
                   [\mathbb{T},v]
  &: \unititl/ \\
\unitetr/ :& [\mathbb{T} \circledast \order{\idiso},
                                                    \langle v,\idiso]
  & \leftrightblackspoon &
                   [\mathbb{T},v]
  &: \unititr/ \\
\swapt/ :& [\mathbb{T}_1 \circledast \mathbb{T}_2,
                                                   \langle v_1,v_2 \rangle]
 & \leftrightblackspoon &
                  [\mathbb{T}_2 \circledast \mathbb{T}_1,
                                                  \langle v_2,v_1 \rangle]
  &: \swapt/ \\
  \assoclt/ :& [\mathbb{T}_1 \circledast (\mathbb{T}_2 \circledast \mathbb{T}_3),
                                        \langle v_1,\langle v_2,v_3 \rangle]
  & \leftrightblackspoon &
                      [(\mathbb{T}_1 \circledast \mathbb{T}_2) \circledast \mathbb{T}_3,
                                       \langle \langle v_1,v_2 \rangle , v_3 \rangle]
  &: \assocrt/ \\
\\
\eta- :& [\order{\idiso},\idiso] & \leftrightblackspoon &
  \forall{k}.[\iorder{c} \circledast \order{c}, \langle 1/c^k, c^k \rangle] \\
\epsilon-: & \exists{k}.[\iorder{c} \circledast \order{c}, \langle 1/c^k, c^k \rangle]
  & \leftrightblackspoon & [\order{\idiso},\idiso] \\
\eta+ :& [\order{\idiso},\idiso] & \leftrightblackspoon &
  \forall{k}.[\order{c} \circledast \iorder{c}, \langle c^k,1/c^k \rangle] \\
\epsilon+ : & \exists{k}.[\order{c} \circledast \iorder{c}, \langle c^k,1/c^k \rangle]
    & \leftrightblackspoon & [\order{\idiso},\idiso]
\end{array}\]
\[\begin{array}{c}
\Rule{}
{}
{\jdg{}{}{\idiso/ :[\mathbb{T},v] \leftrightblackspoon
 [\mathbb{T},v]}}
{}
~~
\Rule{}
{\jdg{}{}{\alpha : c_1 \isotwo c_2}}
{\jdg{}{}{(\order{\alpha})_k :
 [\order{c_1},c_1^k]  \leftrightblackspoon
 [\order{c_2}.c_2^k]}}
{}
\\
\\
\Rule{}
{\jdg{}{}{\rho_1 :
  [\mathbb{T}_1,v_1] \leftrightblackspoon [\mathbb{T}_2,v_2]} \quad
  \rho_2 :
  [\mathbb{T}_2,v_2] \leftrightblackspoon [\mathbb{T}_3,v_3]}
{\jdg{}{}{\rho_1 \circledcirc \rho_2 :
 [\mathbb{T}_1,v_1] \leftrightblackspoon [\mathbb{T}_3},v_3]}
{}
\\
\\
\Rule{}
{\jdg{}{}{\rho_1 :
 [\mathbb{T}_1,v_1] \leftrightblackspoon
 [\mathbb{T}_2,v_2]}
 \qquad \rho_2 :
 [\mathbb{T}_3,v_3] \leftrightblackspoon
 [\mathbb{T}_4,v_4]}
{\jdg{}{}{\rho_1 \circledast \rho_2 :
  [\mathbb{T}_1 \circledast \mathbb{T}_3, \langle v_1,v_3 \rangle]
  \leftrightblackspoon
  [\mathbb{T}_2 \circledast \mathbb{T}_4}, \langle v_2,v_4 \rangle]}
{}
% \\
% \\
% \Rule{}
% {\jdg{}{}{}}
% {\jdg{}{}{\synch_i : \forall{\overline{k}k'}.[\mathbb{T},v]
%   \leftrightblackspoon
%    \forall{\overline{k}}.[\mathbb{T},v[k'\mapsto i]]}}
% {}
\end{array}\]
Each combinator $\rho_\bullet$ has an inverse.
}
\caption{$\Pi^/$ fractional combinators (dependent version)}
\label{pifracd:comb}
\end{figure}

%%%%%
\subsection{Dependent Pointed Types}

@@

\begin{definition}[Dependent $\Pi^/$]
\label{def:pifracd}
The syntax of dependent $\Pi^/$ differs from Def.~\ref{def:pifrac} as follows:
\[\begin{array}{lrcl}
(\textrm{Fractional Types}) & \mathbb{T_\bullet} &::=&
  [\order{c},c^k] \alt
  [\iorder{c},1/c^k] \alt
  [\mathbb{T}_1 \circledast \mathbb{T}_2,\langle v_1,v_2 \rangle] \alt
  \forall k.\mathbb{T_\bullet} \alt \exists k.\mathbb{T_\bullet} \\
 (\textrm{/-combinators}) & \rho_\bullet &:&
  \mathbb{T}_{\bullet 1} \leftrightblackspoon \mathbb{T}_{\bullet 2} ~::=
  [\textit{see Fig.~\ref{pifracd:comb}}]
\end{array}\]
\end{definition}

The first step is to enrich the types to pointed types $\mathbb{T_\bullet}$
where each type $\mathbb{T}$ refers to a particular value $v$ in
$\mathbb{T}$. We then allow indices of values $k$ to be range over a collection
of types. In particular, the type
$\forall k.[\order{c} \circledast \iorder{c}, \langle c^k, 1/c^k \rangle]$
imposes a constraint on the two values associated with the types $\order{c}$ and
$\iorder{c}$; they must be ``synchronized'' sharing the same $k$.

Here is how the circuit in Fig.~\ref{fig:zigzag} evaluates in the forward and
backwards directions:
\[\begin{array}{rcl}
[\order{c},c^i]
&\stackrel{\unitetl/}{\longmapsto}&
  [\order{id}\circledast\order{c}, \langle \idiso,c^i \rangle] \\
&\stackrel{\eta+\circledast\idiso/}{\longmapsto}&
  \forall{k}.[(\order{c}\circledast\iorder{c})\circledast\order{c},
  \langle \langle c^k,1/c^k \rangle,c^i \rangle] \\
&\stackrel{\assocrt/}{\longmapsto}&
  \forall{k}.[\order{c}\circledast(\iorder{c}\circledast\order{c}),
  \langle c^k, \langle 1/c^k,c^i \rangle\rangle] \\
%% &\stackrel{\synch_i}{\longmapsto}&
&k\mapsto i&
  [\order{c}\circledast(\iorder{c}\circledast\order{c}),
  \langle c^i, \langle 1/c^i,c^i \rangle\rangle] \\
&\exists&
  \exists{k}.[\order{c}\circledast(\iorder{c}\circledast\order{c}),
  \langle c^i, \langle 1/c^k,c^k \rangle\rangle] \\
&\stackrel{\idiso/\circledast\epsilon-}{\longmapsto}&
  [\order{c}\circledast\order{\idiso},
  \langle c^i, \idiso \rangle] \\
&\stackrel{\unitetr/}{\longmapsto}&
  [\order{c}, c^i]
\end{array}\]

\[\begin{array}{rcl}
  [\order{c}, c^i]
&\stackrel{\unitetr/}{\longmapsfrom}&
  [\order{c}\circledast\order{\idiso},
  \langle c^i, \idiso \rangle] \\
&\stackrel{\idiso/\circledast\epsilon-}{\longmapsfrom}&
  \forall{k}.[\order{c}\circledast(\iorder{c}\circledast\order{c}),
  \langle c^i, \langle 1/c^k,c^k \rangle\rangle] \\
&\stackrel{\assocrt/}{\longmapsfrom}&
  \forall{k}.[(\order{c}\circledast\iorder{c})\circledast\order{c},
  \langle \langle c^i,1/c^k \rangle,c^k \rangle] \\
&k\mapsto i&
  [(\order{c}\circledast\iorder{c})\circledast\order{c},
  \langle \langle c^i,1/c^i \rangle,c^i \rangle] \\
&\exists&
  \exists{k}.[(\order{c}\circledast\iorder{c})\circledast\order{c},
  \langle \langle c^k,1/c^k \rangle,c^i \rangle] \\
&\stackrel{\eta+\circledast\idiso/}{\longmapsfrom}&
  [\order{\idiso}\circledast\order{c}, \langle \idiso,c^i\rangle] \\
&\stackrel{\unitetl/}{\longmapsfrom}&
  [\order{c},c^i]
\end{array}\]
